---
layout: post
comments: true
title:  "Retrospective survival analysis of different treatments for DIPG"
date:   2016-06-14 16:03:45 -0300
categories: metodologia
ref: retro
lang: en
permalink: /retrospective/
doi: 10.6084/m9.figshare.3489959
tags: pediatric cancer, brain tumors, diffuse intrinsic pontine glioma, clinical trial, retrospective analysis, open science, jekyll, knitr
---

## Retrospective analysis survival curve:

Calculations shown here use data from patients diagnosed with DIPG between 2000-2013, with follow-up until 2014. Follow-up time of the patients in this database is a bit longer than that from the original data used to design VALKYRIE project, hence there are some numerical differences. This post illustrate how to present clinical research data in a transparent and fully reproducible way to an audience.   Including individual patient data (de-identified) as well as the _script_ used to perform statistical analysis of data, this is an example of the possibilities of _open lab notebook_ and _open science_ paradigm. When the prospective trial data are collected, they will be equally published in the same format, becoming permanently available to analysis and criticism by interested third parties. I discussed the inspiration to this approach in this [post][jekyll-migrando] in a personal blog.

This post uses package [_knitr_][knitr] to calculate results at the same time of html parsing with [_jekyll_][jekyll].
I used the excellent example of [_yihui_][yihui] as guide. To format tables with more than that available in knitr, I used package [pander][pander], as very well depicted [here][pander-knitr]. I only made sure to pass the ```type = 'rmarkdown' ``` option to pander.

```{r warning=F, message=F}
require(pander)  
require(knitr)
require(RCurl)
require(survival)
dipg<-read.csv(text = getURL("https://raw.githubusercontent.com/fhcflx/valkyrie/gh-pages/_data/dipg.csv"))
attach(dipg)  
fit1 = survfit(Surv(os,status)~1)  
fit2 = survfit(Surv(os,ifelse(status<1,1,0))~1)  
fit3 = survfit(Surv(os,status)~ifelse(chemo<10,0,1))  
```

```{r Sobrevida, fig.path='figure/source/retrospectiva/', fig.height=10, fig.width=12, fig.cap='Survival of patients with DIPG'}
plot(fit1,xlab="Meses")
```

Patients diagnosed with DIPG in our service between 2000 and 2013. Graph shows survival probability estimate, Kaplan-Meier estimate method, R programming language, package *survival*. Package *RCurl* was additionally used to fecth data from a [.csv][csv] file in github repository. Graph also shows 95% confidence interval.
This is the overall survival and 12-month survival summary:


```{r echo=F}
panderOptions('table.split.table', Inf)
pander(fit1,style = 'rmarkdown')
```

```{r echo=F}
summary(fit1,time=12)
```

Median overall survival was `r round(summary(fit1)$table[7],1)` months and survival at 12 months was `r round(summary(fit1,time=12)$surv,3)*100` %.

This is the _follow-up_ variable summary:


```{r echo=F}
pander(fit2,style = 'rmarkdown')
```

Follow-up median was `r round(summary(fit2)$table[7],1)` monrhs.

Here, graph comparing patients that were treated _as per_ HIT protocols _versus_ other patients:


```{r Sobrevida2, fig.path='figure/source/retrospectiva/', fig.height=10, fig.width=12, fig.cap='Sobrevida de pacientes com DIPG, divididos por tratamento'}
plot(fit3,xlab="Meses")
```

Summary of overall survival and 12-month survival in both groups:

```{r echo=F}
pander(fit3,style = 'rmarkdown')
```
```{r echo=F}
summary(fit3,time=12)
```

12-month survival was `r round(summary(fit3,time=12)$surv[2],3)*100` % for patients treated _as per_ HIT (n = `r summary(fit3,time=12)$n[2]` ) and `r round(summary(fit3,time=12)$surv[1],3)*100` % for the others (n = `r summary(fit3,time=12)$n[1]` ).

Comparing both groups with a non-parametric test that uses Kaplan-Meier estimate, log-rank or Mantel-Haenszel test (see [_survival_][survival] package documentation).

```{r echo=F}
pander(survdiff(Surv(os,status)~ifelse(chemo<10,0,1),dipg),style = 'rmarkdown')
```

Comparing both groups with a semiparametric test, Cox proportional risks model (see [_survival_][survival] package documentation).

```{r echo=F}
pander(coxph(Surv(os,status)~ifelse(chemo<10,0,1),dipg),style = 'rmarkdown')
summary(coxph(Surv(os,status)~ifelse(chemo<10,0,1),dipg))
```

This result indicates that there may be a possible difference in survival when patients are treated _as per_ HIT protocols (with valproic acid).

### References:

- [Jekyll][jekyll]
- [Jekyll, Migrando meu blog para Github Pages usando Jekyll][jekyll-migrando]
- [Knitr, Elegant, flexible and fast dynamic report generation with R][knitr]
- [Knitr, Github][yihui]
- [pander: An R Pandoc Writer][pander]
- [pander, Using pander with kniter][pander-knitr]
- [Package 'survival'][survival]
- [DIPG database][csv]

[jekyll]: https://jekyllrb.com
[jekyll-migrando]: /pharmakon/jekyll/update/2016/05/01/Migrando-meu-blog-para-Github-Pages-usando-Jekyll.html
[knitr]: http://yihui.name/knitr/
[yihui]: https://github.com/yihui/knitr
[pander]: http://rapporter.github.io/pander/
[pander-knitr]: http://rapporter.github.io/pander/knitr.html
[survival]: https://cran.r-project.org/web/packages/survival/survival.pdf
[csv]:https://github.com/fhcflx/valkyrie/blob/master/data/
