---
layout: post
comments: true
title:  "Comparative group from a Brazilian DIPG clinical trial"
date:   2016-06-16 16:03:45 -0300
categories: metodologia
ref: previous
lang: en
permalink: /previous-trial/
doi: 10.6084/m9.figshare.3749790
tags: pediatric cancer, brain tumors, diffuse intrinsic pontine glioma, clinical trial, historical control, project valkyrie
---

## Individual patient data:

In addition to the historical control group of our service, we are to use the information published in a Brazilian clinical study [published][broniscer] in 2000.

```{r warning=F, message=F, echo=F}
require(pander)  
require(knitr)
require(RCurl)
require(survival)
bron<-read.csv(text = getURL("https://raw.githubusercontent.com/fhcflx/valkyrie/gh-pages/_data/broniscer2000table1.csv"))
dipg<-read.csv(text = getURL("https://raw.githubusercontent.com/fhcflx/valkyrie/gh-pages/_data/dipg.csv"))
bron$cens=c(1,1,1,1,1,1,1,1,1,1,0,1,1,1,0,1,1,1,1,1,0,0,0,1,1,1,0,0,0)
fit = survfit(Surv(Survival..days./30.5,cens)~1,bron)
fit.b = survfit(Surv(Survival..days./30.5,ifelse(cens<1,1,0))~1,bron)
fit2 = survfit(Surv(os,status)~1,dipg,subset=(rt>4000&chemo!=10))
```

```{r Sobrevida, fig.path='figure/source/estudo-anterior/', fig.height=10, fig.width=12, fig.cap='Survival of patients with DIPG in the Broniscer _et al_ trial'}
plot(fit,xlab="Months")
```

Group of patients with DIPG (symptoms for less than 6 months, involvement of more than 50% of a pontine segment or two or more segments), treated with radiation (54 to 60 Gy) and tamoxifen during and after RT (200mg/m<sup>2</sup>) at the AC Camargo Hospital in Sao Paulo. The graph shows the estimated probability of survival calculated by the Kaplan-Meier method, using the R programming language, *survival* package. The package *RCurl* was used to capture the data from the .csv file stored on github repository. The graph also shows the 95% confidence interval.

_Reprinted with permission. ©(2016) American Society of Clinical Oncology. All rights reserved. The authors, editors, and ASCO are not responsible for errors or omissions in the translation._


This is the summary of overall survival, in addition to survival in 12 months:

```{r echo=F}
panderOptions('table.split.table', Inf)
pander(fit,style = 'rmarkdown')
```

```{r echo=F}
summary(fit,time=12)
```

Showing a median survival of `r round(summary(fit)$table[7],1)` months and 12-month survival of `r round(summary(fit,time=12)$surv,3)*100`%.

This is the summary of follow-up time:

```{r echo=F}
pander(fit.b,style = 'rmarkdown')
```

Showing a median follow-up of `r round(summary(fit.b)$table[7],1)` months.

Here, the comparison chart between the patients in the historical control group of our service and the patients from the Broniscer _et al_ trial:


```{r Sobrevida2,  fig.path='figure/source/estudo-anterior/', fig.height=10, fig.width=12, fig.cap='Survival of patients with DIPG'}
plot(fit2,xlab="Months",col=4)
lines(fit,col=2)
```

And the summary of overall survival of the two groups, with survival at 12 months:

```{r echo=F}
pander(fit,style = 'rmarkdown')
pander(fit2,style = 'rmarkdown')
```
```{r echo=F}
summary(fit,time=12)
summary(fit2,time=12)
```

Showing a survival at 12 months of `r round(summary(fit,time=12)$surv[1],3)*100` % for patients in the Broniscer _et al_ trial (n = 29) and `r summary(fit,time=12)$n[1]`% in the historical control group of our service (n = `r summary(fit2,time=12)$n[1]`).

Permission to use the data from the main table from [Broniscer _et al_][broniscer] was granted to us by [ASCO][asco] in 2016. These data are protected by copyright and should not be reused without permission from rightholder.

### References:

- [Broniscer A, Leite CC, Lanchote VL, Machado TM, Cristófani LM. Radiation therapy and high-dose tamoxifen in the treatment of patients with diffuse brainstem gliomas: results of a Brazilian cooperative study. Brainstem Glioma Cooperative Group. J Clin Oncol. 2000 Mar;18(6):1246-53.][broniscer]
- [ASCO: American Society of Clinical Oncology][asco]

[broniscer]: http://jco.ascopubs.org/content/18/6/1246.long
[asco]: http://www.asco.org
